A fully labelled proof system for intuitionistic modal logics

### 2019-10-07 caleidoscope/introduction.tex
proof read section 1, 2 and 3

Since their introduction in the 1980s by Gabbay~\cite{gabbay:96},
\emph{labelled proof calculi} have been widely used by proof theorist
to give sound, complete and cut-free deductive systems to a broad range
of logics. Unlike so-called \emph{internal calculi}, like
hypersequents~\cite{avron:elc96}, nested
sequents~\cite{kashima:sl94,brunnler:aml09,poggiolesi:tmp09}, 2-sequents~\cite{masini:apal92}, or linear nested sequents~\cite{lellmann?},
labelled calculi have the advantage of being more uniform and being able to 
accommode a larger class of logics.

@@

However,
in this paper we are concerned with \emph{intuitionistic modal
  logics}, whose Kripke semantics is based on \emph{birelational}
frames, i.e., they have two binary relations instead of one: one
relation $R$ that corresponds to the accessibility relation in 
Kripke frames for modal logics, and a relation $\le$ that
corresponds to the preorder relation in Kripke frames for
intuitionistic logic. Consequently, standard labelled systems for
these logics have certain shortcomings:
\begin{enumerate}
\item The transitivity rule in~\eqref{eq:Rtrans} corresponds to both
  versions of the $\fourax$-axiom
  \begin{equation}
    \label{eq:4ax}
    \fm{\BOX A\IMP \BOX\BOX A}
    \qquand
    \fm{\DIA\DIA A\IMP\DIA A}
    \quadcm
  \end{equation}
  as they are equivalent in classical
  modal logic. However, in intuitionistic modal logic they are not
  equivalent, and even though the logic $\IKfour$, i.e., the
  intuitionistic version of the modal logic $\Kfour$, contains both
  axioms, they can also be added independently to the logic $\IK$ (an
  intuitionistic version of $\K$). However, the proof theory of these
  new logics has not been studied before, as no existing labelled (or
  label-free) proof system can handle them, even though the
  corresponding frame conditions
  \begin{equation}
    \label{eq:4frame}
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
    \quand
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
    \;,
  \end{equation}
  \todo{add graphical representation of the conditions}
  respectively, have already been studied
  in~\cite{plotkin:stirling:86}.
  \sonia{the general condition is correct but not the application to 4}
  \item The correspondence between the syntax and the semantics is not
  as clean as one would expect. As only the $R$-relation (and not the
  $\le$-relation) of the frame is visible in an ordinary labelled
  sequent, we only have that a sequent $\Gamma$ is provable if and
  only if is is valid in all \emph{graph-consistent}\footnote{This
    means that every layer in the model can be lifted to any future of
    any world in that layer. See~\cite{simpson:phd} and~\cite{mar:str:tableaux17} for a formal definition and discussion.}   
  models, 
  \sonia{this is only true for logics outside of the path axioms.}
  \todo{check that Simpson's system is sound and complete for all logics defined by path axioms. Is it that the axiomatisation is also sound and complete for every models and not only graph-consistent ones?}
  as already observed in by Simpson in his PhD thesis~\cite{simpson:phd}
  and considered as an inelegant solution (see also \cite{mar:str:tableaux17}).
\end{enumerate}
In order to address these two concerns we propose here to enrich usual 
labelled sequent by allowing both, relational atoms of the
form $\futs xy$ and of the form $\accs xy$. Consequently, we can easily translate the frame
conditions in~\eqref{eq:4frame} into inference rules:
\begin{equation}
  \label{eq:4rules}
  \vlinf{\rn{4^{\BOX}}}{\proviso{$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
     \B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
  \qquand
  \vlinf{\rn{4^{\DIA}}}{\proviso{$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
\end{equation}
This allows us to define cut-free deductive systems for a wide range
of logics that could not be treated before.  Furthermore, the relation
between syntax and semantics is as one would expect: A sequent is
provable in our system if and only if it is valid in all models.

Besides that, there is another pleasant observation to make about our
system. Ordinary labelled sequent systems for intuitionistic modal
logic are single-conclusion~\cite{simpson:phd}. 
\sonia{No! the system for epistemic logic by Maffezioli, Naibo and Negri~\cite{maffezioli:etal:synthese13} is multi-conclusion}
%
The same is true for the corresponding nested sequent
systems~\cite{str:fossacs13,marin:str:aiml}. It is possible to express
Maehara style multiple-conclusion systems in nested
sequents~\cite{kuz:str:maehara}, and therefore also in ordinary
labelled sequents. However, also in these systems there are rules
($\rlabrn\IMP$ and $\rlabrn\BOX$) that force a single-conclusion
premise, even though this is not the case in labelled
systems~\cite{negri:jpl2005} or nested sequents~\cite{fitting:83} for
intuitionistic logic. Our system eliminates this
discrepancy. Consequently, every rule in our system is invertible, as
we never delete information.

>>>>

Since their introduction in the 1980s by Gabbay~\cite{gabbay:96},
\emph{labelled proof calculi} have been widely used by proof theorist
to give sound, complete and cut-free deductive systems to a broad range
of logics. Unlike so-called \emph{internal calculi}, like
hypersequents~\cite{avron:elc96}, nested
sequents~\cite{kashima:sl94,brunnler:aml09,poggiolesi:tmp09}, 2-sequents~\cite{masini:apal92}, or linear nested sequents~\cite{lellmann:tableaux15},
labelled calculi have the advantage of being more uniform and being able to 
accommode a larger class of logics.

@@

However,
in this paper we are concerned with \emph{intuitionistic modal
  logics}, whose Kripke semantics is based on \emph{birelational}
frames, i.e., they have two binary relations instead of one: one
relation $R$ that corresponds to the accessibility relation in 
Kripke frames for modal logics, and a relation $\le$ that
corresponds to the preorder relation in Kripke frames for
intuitionistic logic. Consequently, standard labelled systems for
these logics have certain shortcomings:
\begin{enumerate}
\item The transitivity rule in~\eqref{eq:Rtrans} can be axiomatised by the conjunction of the two
  versions of the $\fourax$-axiom
  \begin{equation}
    \label{eq:4ax}
    \fm{\BOX A\IMP \BOX\BOX A}
    \qquand
    \fm{\DIA\DIA A\IMP\DIA A}
    \quadcm
  \end{equation}
  which are equivalent in classical
  modal logic. However, in intuitionistic modal logic they are not
  equivalent, and even though the logic $\IKfour$, i.e., the
  intuitionistic version of the modal logic $\Kfour$, contains both
  axioms, they can also be added independently to the logic $\IK$ (an
  intuitionistic version of $\K$). The proof theory of these
  new logics has not been studied before; no existing labelled (or
  label-free) proof system can handle them, even though the
  corresponding frame conditions
  \begin{equation}
    \label{eq:4frame}
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
    \quand
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
    \;,
  \end{equation}
%  \todo{add graphical representation of the conditions}
  $$
  \xymatrix{
  	x' \ar@{.>}[drr]^R \\
  	x \ar@{.>}[u]^\le \ar@{->}[r]_{R} & y \ar@{->}[r]_{R} & z
  }
  \qquad\qquad
  \xymatrix{
  	&& z''  \\
  	x  \ar@{.>}[urr]^R \ar@{->}[r]_{R} & y \ar@{->}[r]_{R} & z \ar@{.>}[u]^\le
  }
  $$
  respectively, have already been studied
  in~\cite{plotkin:stirling:86}.
%  \sonia{the general condition is correct but not the application to 4}
  \item The correspondence between the syntax and the semantics is not
  as clean as one would expect. As only the $R$-relation (and not the
  $\le$-relation) of the frame is visible in an ordinary labelled
  sequent, we only have that a sequent $\Gamma$ is provable if and
  only if is is valid in all \emph{graph-consistent}\footnote{This
    means that every layer in the model can be lifted to any future of
    any world in that layer. See~\cite{simpson:phd} and~\cite{mar:str:tableaux17} for a formal definition and discussion.}   
  models, 
  \sonia{this is only true for logics outside of the path axioms.}
  \todo{check that Simpson's system is sound and complete for all logics defined by path axioms. Is it that the axiomatisation is also sound and complete for every models and not only graph-consistent ones?}
  \sonia{yes, they are sound and complete for the general models, but they do not \emph{characterize} the frames, they are not \emph{frame conditions}, as opposed to the \emph{correspondence result} of Plotkin and Sterling.}
  as already observed in by Simpson in his PhD thesis~\cite{simpson:phd}
  and considered as an inelegant solution (see also \cite{mar:str:tableaux17}).
\end{enumerate}
In order to address these two concerns we propose here to enrich usual 
labelled sequent by allowing both, relational atoms of the
form $\futs xy$ and of the form $\accs xy$. Consequently, we can easily translate the frame
conditions in~\eqref{eq:4frame} into inference rules:
\begin{equation}
  \label{eq:4rules}
  \vlinf{\rn{4^{\BOX}}}{\proviso{$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
     \B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
  \qquand
  \vlinf{\rn{4^{\DIA}}}{\proviso{$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
\end{equation}
This allows us to define cut-free deductive systems for a wide range
of logics that could not be treated before.  Furthermore, the relation
between syntax and semantics is as one would expect: A sequent is
provable in our system if and only if it is valid in all models.

Besides that, there is another pleasant observation to make about our
system. Ordinary labelled sequent systems for intuitionistic modal
logic are single-conclusion~\cite{simpson:phd}. 
\sonia{No! the system for epistemic logic by Maffezioli, Naibo and Negri~\cite{maffezioli:etal:synthese13} is multi-conclusion}
\todo{need to say that it is the same principle as in their work, but we use in a more general setting and extend it to a framework for many intuitionistic modal logics while keeping the nice invertibility property.}
%
The same is true for the corresponding nested sequent
systems~\cite{str:fossacs13,marin:str:aiml}. It is possible to express
Maehara style multiple-conclusion systems in nested
sequents~\cite{kuz:str:maehara}, and therefore also in ordinary
labelled sequents. However, also in these systems there are rules
($\rlabrn\IMP$ and $\rlabrn\BOX$) that force a single-conclusion
premise, even though this is not the case in labelled
systems~\cite{negri:jpl2005} or nested sequents~\cite{fitting:83} for
intuitionistic logic. Our system eliminates this
discrepancy. Consequently, every rule in our system is invertible, as
we never delete information.
